Significant wind disturbances induced by giant dunes.

### 2021-12-15 Paper/main.tex
Rewriting conclusion

 The comparison of local and regional wind data gives a direct evidence of the giant dunes feedback on the flow. In flat areas, the matching between both datasets highlights the ability of the latest generation of climate reanalysis to predict the wind flow up to scales $\sim 10~\textrm{km}$, i.e the grid model. When smaller scale topographies are present (giant dunes in our case), locally measured wind regimes may significantly differ from the regional ones. Furthermore, we link these disturbances induced by the dunes to their interaction with the lower part of the atmospheric vertical structure, and more specifically to its circadian variability. During the day, the top of the ABL is high enough to limit the interaction of the capping layer and the FA stratification with the giant dunes, resulting in a low flow confinement, and thus small wind disturbances. During the night, the small ABL height induces a stronger flow confinement, associated with large wind deviation and acceleration or deceleration. Interestingly, we also found that this effect could be counterbalanced by the presence of large wind velocities, capable of deforming the capping layer and/or the FA stratification and thus decreasing the confinement effect.

 Simple linear model such as the one of \citet{andreotti2009} also suggests that larger wind disturbances occur under strong flow confinement such as described above. However, they are unable to reproduce the magnitude of the observed deviations, probably due to the presence of hydrodynamical non-linear effects, negligible in low confinement situations, but not otherwise (see Fig.~\ref{Fig12_supp} and Appendix 1). Another limit in the comparison between theoretical predictions and measured is induces by the the single-point measurements. To have reliable representations of the flow structures related to wind disturbances, additional measurements in different places on and near the same topographical obstacle are needed.

 This study highlights the interaction between giant dunes and the atmospheric boundary layer, thus supporting for example the way the capping layer acts as a bounding surface limiting dune growth \citep{andreotti2009, gunn2021}. This interaction also have implications at smaller scales, where bedforms then develop from the disturbed wind instead of the regional one. Differences between larger and smaller scale (thus older and more recent) dune patterns are observed ubiquitously, and have sometimes in the literature been attributed to climatic changes in wind regimes \citep{refs}. Here, we suggest using this feedback mechanism that current winds can explain dune patterns at all scales, such as the linear dunes ($\sim 50$ m -wide) elongating within the interdune between two giant linear dunes ($\sim 2$ km -wide) in the Namib Sand Sea (see Fig.~\ref{Fig6}).

>>>>

 The comparison of local (direct measurements) and regional (climate reanalysis) wind data reveals the giant dunes feedback on the flow. In flat areas, the matching between both datasets confirms the ability of the ERA5Land climate reanalysis to predict the wind flow down to scales $\sim 10~\textrm{km}$, i.e the grid model. When smaller scale topographies are present (giant dunes in our case), locally measured wind regimes may significantly differ from the regional ones. Furthermore, we link these disturbances induced by the dunes to their interaction with the lower part of the atmospheric vertical structure, and more specifically to its circadian variability. During the night, the presence of a shallow atmospheric boundary layer (ABL) induces a strong flow confinement, associated with large wind deviation and acceleration or deceleration. During the day, the capping layer is high enough to prevent its interaction with the giant dunes, resulting in a low flow confinement, and thus smaller wind disturbances. Interestingly, we also found that this effect could be counterbalanced by the presence of large wind velocities, capable of deforming the capping layer and/or the FA stratification, thus decreasing the confinement effect.

 Simple linear models such as the one of \citet{andreotti2009} also suggests that larger wind disturbances occur under strong flow confinement such as described above. However, they are unable to reproduce the magnitude of the observed deviations, probably due to the presence of hydrodynamical non-linear effects, negligible in low confinement situations, but not otherwise (see Fig.~\ref{Fig12_supp} and Appendix 1). Additionally, note that the spatial flow structures associated with the dune feedback on the wind can not be studied with our single-point measurements. Measurements in different places on and near the same topographical obstacle are then needed for further comparisons with models.

 This study highlights the interaction between giant dunes and the atmospheric boundary layer, supporting as well the way the capping layer acts as a bounding surface limiting dune growth \citep{andreotti2009, gunn2021}. This interaction also have implications at smaller scales, where bedforms then develop from the wind disturbed by the giant dunes, instead of the regional wind regime. Differences between larger and smaller scale dune patterns are observed ubiquitously. As larger dunes are often older than smaller ones, sometimes this discrepancy have been previously attributed to climatic changes in wind regimes \citep{refs}. Here, using this feedback mechanism, we suggest that current winds are also able to explain dune patterns at all scales. For example, this seems to be the case for the linear dunes ($\sim 50$ m -wide) elongating within the interdune between two giant linear dunes ($\sim 2$ km -wide) in the Namib Sand Sea (see Fig.~\ref{Fig6}).
